Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-04 doc/sc2012/legion.tex
first two sections

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  Because the
costs of communication within these architectures vary by several orders of magnitude, 
the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
to achieve good performance it is crucial that the programmer
and the compiler be able to reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  
Most contemporary
programming systems have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.  
%
% FIXME
% Sean would like a bit more on why other systems don't support pointer data structures (e.g., the difficulty of
% alias analysis.  I thought about it and feel it is too much of a digression: mentioning that in the first paragraph
% makes it sound like it will be an important part of the paper.
%
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data
and to make explicit relationships useful for reasoning about locality
and independence.
A logical region names a set of objects.  Logical regions are first-class 
values in Legion and may be dynamically allocated, deleted and stored
in data structures. Regions can also be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into (usually) disjoint {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} ({\em read-only}, {\em read-write}, and {\em reduce}) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Privileges express how a task may use
its region arguments, providing data dependence information that
is used to guide the extraction of parallelism.  For example, if two tasks
access the same region with read-only privileges the two tasks
can potentially be run in parallel.
Coherence properties express the required semantics of concurrent region accesses.
For example, if the program executes $f_1(r); f_2(r)$ and tasks $f_1$
and $f_2$ both require {\em exclusive} access to region $r$, then Legion guarantees the result will
be as if $f_1(r)$ completes before $f_2(r)$ begins.   On the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either task $f_1(r)$
appears to run entirely before $f_2(r)$ or vice versa.

Legion programs are written using logical regions, which do not commit
to any particular placement of the data in the machine.  At runtime,
each logical reagion has one or more {\em physical instances}.  Often
logical regions correspond to a single physical instance, but it is also common and useful
to have multiple physical copies of a logical region's contents in different parts of the machine 
to duplicate and distribute read-only data, to keep an old copy of a region to break a 
write-after-read dependence, or to allow independent, local reductions that are later combined.

@@

\item We present the results of experiments on three applications:
  fluid-flow on a regular grid, a three-level AMR code solving a
  standard heat diffusion equation and a circuit simulation.
  We run each of these four codes and best available reference
  versions on three different clusters of multicore processors with
  attached GPUs: a small cluster of our own (4 nodes with 12 CPUs and 2 GPUs per node), a medium-sized cluster
  designed for visualization applications (10 nodes with 12 CPUs and 5 GPUs per node), and up to a third of the
  Keeneland supercomputer (32 nodes, 12 CPUs and 3 GPUs per node).
%
% FIXME Give the same numbers as in the abstract.
%

>>>>

Modern parallel machines are increasingly complex, with deep,
distributed memory hierarchies and heterogeneous processing units.  Because the
costs of communication within these architectures vary by several orders of magnitude, 
the penalty for mistakes in the placement of data or computation is usually very poor performance.  Thus,
to achieve good performance the programmer
and the programming system must reason about {\em locality} (data resident
close to computation that uses it) and {\em independence} (computations
operating on disjoint data, and therefore not requiring communication and able to be placed in possibly distant parts of the machine).  
Most contemporary
programming systems have no facilities for the programmer to express locality
and independence.  The few languages that do focus primarily on
array-based locality \cite{Fatahalian06,CHAPEL04,UPC99} and 
avoid irregular pointer data structures.  
%
% FIXME
% Sean would like a bit more on why other systems don't support pointer data structures (e.g., the difficulty of
% alias analysis.  I thought about it and feel it is too much of a digression: mentioning that in the first paragraph
% makes it sound like it will be an important part of the paper.
%
%encode such structures in arrays, which 
%data structures encoded in arrays which are impossible for compilers
%and runtime systems to reason about with regard to locality and independence.

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data
and to make explicit relationships useful for reasoning about locality
and independence.
A logical region names a set of objects.  Logical regions are first-class 
values in Legion and may be dynamically allocated, deleted and stored
in data structures. Regions can also be passed as arguments
to distinguished functions called {\em tasks} that access the data in
those regions, providing locality information.  Logical regions may be
{\em partitioned} into disjoint or aliased (overlapping) {\em subregions}, providing
information for determining independence of computations.
Furthermore, computations access logical regions with particular {\em
  privileges} ({\em read-only}, {\em read-write}, and {\em reduce}) and {\em coherence} (e.g., {\em exclusive access} and {\em
  atomic access}, among others).  Privileges express how a task may use
its region arguments, providing data dependence information that
is used to guide the extraction of parallelism.  For example, if two tasks
access the same region with read-only privileges the two tasks
can potentially be run in parallel.
Coherence properties express the required semantics of concurrent region accesses.
For example, if the program executes $f_1(r); f_2(r)$ and tasks $f_1$
and $f_2$ both require {\em exclusive} access to region $r$, then Legion guarantees the result will
be as if $f_1(r)$ completes before $f_2(r)$ begins.   On the other hand, if
the tasks access $r$ with {\em atomic} coherence, then Legion
guarantees only atomicity of the tasks with respect to $r$: either task $f_1(r)$
appears to run entirely before $f_2(r)$ or vice versa.

Logical regions do not commit to any particular layout of the data or
placement in the machine.  At runtime, each logical region has one or
more {\em physical instances} assigned to specific memories .  It is often useful to have multiple
physical instances of a logical region (e.g., to duplicate read-only data, or to allow
independent reductions that are later combined).

@@

\item We present results of experiments on three applications:
  fluid-flow on a regular grid, a three-level AMR code solving a
  heat diffusion equation and a circuit simulation.
  We compare each application with best available reference
  versions on three different clusters of multicore processors with
  GPUs, including the Keeneland supercomputer.
%
% FIXME Give the same numbers as in the abstract.
%
